Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-27 analysis/paper/paper.qmd
minor changes

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997]. Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively. Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].

@@

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009). Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

@@

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal elements and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal elements were quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP. Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

# bones with surface modifications
burned_deer <- fauna_deer_only %>% filter(burning == "yes") %>% count(period) %>% rename(burned = n)
cutmark_deer <- fauna_deer_only %>% filter(cutmarks == "yes") %>% count(period) %>% rename(cutmarks = n)
frac_deer <- fauna_deer_only %>% filter(fractures  == "yes") %>%
  filter(`部位/名稱` %in% c("脛骨","股骨","肱骨","橈骨","上肢骨","掌骨","掌骨或蹠骨","蹠骨")) %>%
  count(period) %>% rename(fractures = n)
list_modif <- list(burned_deer, cutmark_deer, frac_deer, deer_NISP)

# join
deer_modify <-
  list_modif %>%
  reduce(inner_join, by = 'period') %>%
  mutate(cut_per = round(cutmarks/NISP, 3) * 100, burn_per = round(burned/NISP, 3) * 100,
         frac_per = round(fractures/NISP, 3) * 100) %>%
  mutate(burned = paste(burned, paste0("(", burn_per, " %)")),
         cutmarks = paste(cutmarks, paste0("(", cut_per, " %)")),
         fractures = paste(fractures, paste0("(", frac_per, " %)"))) %>%
  rename(`cut marks` = cutmarks) %>%
  select(-cut_per, -frac_per, -burn_per)

@@

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023]. In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).

@@

```{r}
#| label: fig-cutmark-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage. Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows. Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."

@@

```{r}
#| label: fig-fracture-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks. Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."

@@

# Discussion

@@

# Conclusion
Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997]. Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively. Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].

@@

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in daily practices can reveal the strategies Indigenous people used to respond to resource explortations by foreign presence. I will examine the zooarchaeological assemblages from the upper culture layer at Kiwulan, a large settlement dating to 14-19th centuries. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as European colonization or intercultural contact [@Defrance2009]. Through a zooarchaeological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

@@

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers were excluded since they were less likely related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yield and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation and quantities compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP. Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, which were present in CL5, increased to the highest number in CL4, and then decreased thereafter. None of cranial bones were found in CL6 and CL2. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

# bones with surface modifications
burned_deer <- fauna_deer_only %>% filter(burning == "yes") %>% count(period) %>% rename(burned = n)
cutmark_deer <- fauna_deer_only %>% filter(cutmarks == "yes") %>% count(period) %>% rename(cutmarks = n)
frac_deer <- fauna_deer_only %>% filter(fractures  == "yes") %>%
  filter(`部位/名稱` %in% c("脛骨","股骨","肱骨","橈骨","上肢骨","掌骨","掌骨或蹠骨","蹠骨")) %>%
  count(period) %>% rename(fractures = n)
post_deposit <-fauna_deer_only %>% filter(!is.na(`保存狀態`)) %>% count(period) %>%
  rename(`post-deposition` = n)

list_modif <- list(burned_deer, cutmark_deer, frac_deer, post_deposit, deer_NISP)

# join
deer_modify <-
  list_modif %>%
  reduce(inner_join, by = 'period') %>%
  mutate(cut_per = round(cutmarks/NISP, 3) * 100, burn_per = round(burned/NISP, 3) * 100,
         frac_per = round(fractures/NISP, 3) * 100, post_per= round(`post-deposition`/NISP, 3) * 100) %>%
  mutate(burned = paste(burned, paste0("(", burn_per, " %)")),
         cutmarks = paste(cutmarks, paste0("(", cut_per, " %)")),
         fractures = paste(fractures, paste0("(", frac_per, " %)")),
         `post-deposition` = paste(`post-deposition`, paste0("(", post_per, " %)"))) %>%
  rename(`cut marks` = cutmarks) %>%
  select(-cut_per, -frac_per, -burn_per, -post_per)

@@

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing. In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).

@@

```{r}
#| label: fig-cutmark-img
#| echo: false
#| fig-cap: "Examples of deer bones with cut marks, chop surfaces, and burning damage. Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows. Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."

@@

```{r}
#| label: fig-fracture-img
#| echo: false
#| fig-cap: "Examples of deer bones with fractures, natural modifications, and chewing marks. Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."

@@

# Discussion
<!--summary of key results-->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption since CL6 until CL2, and then shifted to cattle in CL1. Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes over time. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences relative to foreign presences for the following discussion. The sequences include the pre-European period (CL6, CL5), the European period (CL4), post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and activities in northern Taiwan, including the involvement into the long-distance trade network, part of the tribute system, and experiences of occasional attacks in cases in Yilan.

Faunal remains recovered from the layer associated with the European presence reveals the highest quantity, followed by the pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of boar among faunal remains suggests a slightly greater reliance on boar as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles. A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, boar and munjact remains during the European period. Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period. During the Chinese period, the low quantity of deer in CL2 and the predomination of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigeonus population in this region.

The skeletal parts of deer demonstrate a different

The cutmarks found on bone portions related to skinning correspond to the description of deerskin as a commodity during the maritime era (Chen, 2004; Cheng, 2017; Höllmann, 1991). Interestingly, a similar butchering pattern related to skinning suggests that hides were produced at the same rate during the European period, despite the high demand for deerskin. However, the dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995). Moreover, the double increase in pig bones after the European arrival may suggest an increased reliance on pigs. The shape of the third molars of pigs indicates that most were domesticated. This implies that pigs played a more significant role in the diet after Europeans arrived. Eventually, cows replaced pigs during the Chinese period, which might be a Chinese influence.

By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the

<!--population dynamics and pottery assembalges-->

<!--environmental changes in general and taphonomic issues-->

<!--resilence-->

@@

The European maritime cultural exchange to some extent altered the material culture of the Indigenous peoples in northeastern Taiwan by adopting foreign goods or customs into their daily lives. However, a sign of indigenous resilience in the face of European influences and resource exploitation may be seen in the continuing tradition of pottery production and deer consumption with few changes, even though changes were observed through the uneven distribution of prestige items, such as glass beads and porcelains, across different households in the living area and burials (Wang and Marwick, 2020a; 2021).

The continuous practices of ceramic making and deer consumption by the Indigenous peoples before and after the European arrival can be viewed as a way of negotiating their identities and maintaining their traditions. Everyday practices are essential in connecting history and social memories, which helps maintain tradition in a setting of frequent cultural exchange or asymmetrical relations (Silliman, 2009). Moreover, ceramics could be an item that conveys social meaning through its style, such as decorations, and be used to show group identity (Croucher and Wynne-Jones, 2006; Pikirayi, 2007). It is worth noting that the pottery found at Kiwulan, with its typical round body and incised geometric motifs, is also commonly found across many sites during this period in northeastern Taiwan (Taiwan History, 2005). Similar geometric motifs and vessel shapes in the region might suggest the communication of Indigenous identity within the context of ongoing cross-cultural interactions with outsiders, including Europeans and Chinese. Furthermore, despite the availability of foreign items, such as iron pans that could have potentially replaced local pottery for cooking purposes, local ceramics have been used continuously for over 6,000 years (Chen, 2007). This emphasizes the significant role of pottery-making in the lives of Indigenous communities.

By situating the findings within a pericolonial context (Acabado, 2017), locally made pottery and faunal remains reveal cultural persistence and resilience among Indigenous peoples in northeast Taiwan. The interaction between Europeans and Indigenous peoples in the region appears to have been based on negotiation and bartering (Torrence, 2000). For instance, the Dutch reduced and eventually waived annual tribute in 1648 in response to the resistance to paying tribute by most Indigenous settlements in Yilan (Blussé and Everts, 2000, pp. 534, 536, 558; Kang, 2012). The continuous consumption of deer and consistent skinning patterns before and after the arrival of Europeans indicate a more adaptable tribute payment system, whereby the Indigenous peoples did not need to hunt more deer to meet European demand. However, it is important to note that there may have been a gradual decline in deer numbers since the post-European period, which is supported by the increased number of pigs found in the same period. This could suggest cultural resilience through dietary changes to adapt to deer overhunting. A more comprehensive investigation across different Indigenous settlements in northeastern Taiwan could provide a clearer understanding of overall dietary patterns.

# Conclusion

Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.

Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.
